Visualising How Non-linear Dimension Reduction Warps Your Data

### 2024-03-22 paper.qmd
Added write up for digit 1 MNIST data

plot_list2_pbmc_error <- error_df |>
  mutate(type = factor(type , levels = c("no error", "error 0-5", "error 5-10", "error 10-15", "error 15-20", "error 20-25", "error greter than 25"))) |>
    ggplot(aes(x = UMAP1,
               y = UMAP2, color = type))+
    geom_point(alpha=0.5, size = 0.1) +
    coord_equal() +
    theme(plot.title = element_text(hjust = 0.5, size = 18, face = "bold")) + #ggtitle("(a)") +
  theme_linedraw() +
    theme(plot.title = element_text(size = 7, hjust = 0.5, vjust = -0.5),
              axis.title.x = element_blank(), axis.title.y = element_blank(),
              axis.text.x = element_blank(), axis.ticks.x = element_blank(),
              axis.text.y = element_blank(), axis.ticks.y = element_blank(),
              panel.grid.major = element_blank(), panel.grid.minor = element_blank(), #change legend key width
        legend.title = element_text(size=5), #change legend title font size
        legend.text = element_text(size=4),
         legend.key.height = unit(0.25, 'cm'),
         legend.key.width = unit(0.25, 'cm')) +
  scale_color_manual(values=c("#fee0d2", "#fcbba1",
                                         "#fc9272", "#fb6a4a", "#ef3b2c",
                                         "#cb181d", "#a50f15")) +
  annotate(geom = 'text', label = 'b', x = -Inf, y = Inf, hjust = -0.5, vjust = 1.5, size = 3)
```

@@

The MNIST dataset consists of grayscale images of handwritten digits [@lecun2010]. @Yingfan2021 used this dataset to demonstrate how the PaCMAP preserves local structure. We select the 2D embedding of PaCMAP for the handwritten digit 1. As shown in @fig-pacmapauthor, we observe a nonlinear structure present in 2D space. To answer the following questions,

@@

@@

<!--images with position 1-->
```{r}
#| warning: false
#| echo: false

pixels_gathered <-  mnist_data |>
  mutate(instance = row_number()) |>
  gather(pixel, value, -Label, -instance) |>
  tidyr::extract(pixel, "pixel", "(\\d+)", convert = TRUE) |>
  mutate(pixel = pixel - 2, x = pixel %% 28, y = 28 - pixel %/% 28) |>
  filter(instance %in% c(3710, 2391, 2385, 5030, 6475))

theme_set(theme_light())
imge_error_sample1 <- pixels_gathered |> 
  ggplot(aes(x, y, fill = value)) +  
  geom_tile() +  
  facet_wrap(~ instance, ncol = 5) +
  theme(legend.position = "none", plot.title = element_text(size = 7, hjust = 0.5, vjust = -0.5),
              axis.title.x = element_blank(), axis.title.y = element_blank(),
              axis.text.x = element_blank(), axis.ticks.x = element_blank(),
              axis.text.y = element_blank(), axis.ticks.y = element_blank(),
              panel.grid.major = element_blank(), panel.grid.minor = element_blank(), #change legend key width
        legend.title = element_text(size=5), #change legend title font size
        legend.text = element_text(size=4),
         legend.key.height = unit(0.25, 'cm'),
         legend.key.width = unit(0.25, 'cm'),
        strip.text = element_blank())
```

<!--images with position 2-->
```{r}
#| warning: false
#| echo: false

pixels_gathered <-  mnist_data |>
  mutate(instance = row_number()) |>
  gather(pixel, value, -Label, -instance) |>
  tidyr::extract(pixel, "pixel", "(\\d+)", convert = TRUE) |>
  mutate(pixel = pixel - 2, x = pixel %% 28, y = 28 - pixel %/% 28) |>
  filter(instance %in% c(7679, 3568, 1312, 3311, 1097))

theme_set(theme_light())
imge_error_sample2 <- pixels_gathered |> 
  ggplot(aes(x, y, fill = value)) +  
  geom_tile() +  
  facet_wrap(~ instance, ncol = 5) +
  theme(legend.position = "none", plot.title = element_text(size = 7, hjust = 0.5, vjust = -0.5),
              axis.title.x = element_blank(), axis.title.y = element_blank(),
              axis.text.x = element_blank(), axis.ticks.x = element_blank(),
              axis.text.y = element_blank(), axis.ticks.y = element_blank(),
              panel.grid.major = element_blank(), panel.grid.minor = element_blank(), #change legend key width
        legend.title = element_text(size=5), #change legend title font size
        legend.text = element_text(size=4),
         legend.key.height = unit(0.25, 'cm'),
         legend.key.width = unit(0.25, 'cm'),
        strip.text = element_blank())
```

<!--images with position 3-->
```{r}
#| warning: false
#| echo: false

pixels_gathered <-  mnist_data |>
  mutate(instance = row_number()) |>
  gather(pixel, value, -Label, -instance) |>
  tidyr::extract(pixel, "pixel", "(\\d+)", convert = TRUE) |>
  mutate(pixel = pixel - 2, x = pixel %% 28, y = 28 - pixel %/% 28) |>
  filter(instance %in% c(3552, 7853, 6489, 7689, 6690))

theme_set(theme_light())
imge_error_sample3 <- pixels_gathered |> 
  ggplot(aes(x, y, fill = value)) +  
  geom_tile() +  
  facet_wrap(~ instance, ncol = 5) +
  theme(legend.position = "none", plot.title = element_text(size = 7, hjust = 0.5, vjust = -0.5),
              axis.title.x = element_blank(), axis.title.y = element_blank(),
              axis.text.x = element_blank(), axis.ticks.x = element_blank(),
              axis.text.y = element_blank(), axis.ticks.y = element_blank(),
              panel.grid.major = element_blank(), panel.grid.minor = element_blank(), #change legend key width
        legend.title = element_text(size=5), #change legend title font size
        legend.text = element_text(size=4),
         legend.key.height = unit(0.25, 'cm'),
         legend.key.width = unit(0.25, 'cm'),
        strip.text = element_blank())
```

<!--images with position 4-->
```{r}
#| warning: false
#| echo: false

pixels_gathered <-  mnist_data |>
  mutate(instance = row_number()) |>
  gather(pixel, value, -Label, -instance) |>
  tidyr::extract(pixel, "pixel", "(\\d+)", convert = TRUE) |>
  mutate(pixel = pixel - 2, x = pixel %% 28, y = 28 - pixel %/% 28) |>
  filter(instance %in% c(1380, 6057, 2347, 5946, 3355))

theme_set(theme_light())
imge_error_sample4 <- pixels_gathered |> 
  ggplot(aes(x, y, fill = value)) +  
  geom_tile() +  
  facet_wrap(~ instance, ncol = 5) +
  theme(legend.position = "none", plot.title = element_text(size = 7, hjust = 0.5, vjust = -0.5),
              axis.title.x = element_blank(), axis.title.y = element_blank(),
              axis.text.x = element_blank(), axis.ticks.x = element_blank(),
              axis.text.y = element_blank(), axis.ticks.y = element_blank(),
              panel.grid.major = element_blank(), panel.grid.minor = element_blank(), #change legend key width
        legend.title = element_text(size=5), #change legend title font size
        legend.text = element_text(size=4),
         legend.key.height = unit(0.25, 'cm'),
         legend.key.width = unit(0.25, 'cm'),
        strip.text = element_blank())
```

<!--images with position 5-->
```{r}
#| warning: false
#| echo: false

pixels_gathered <-  mnist_data |>
  mutate(instance = row_number()) |>
  gather(pixel, value, -Label, -instance) |>
  tidyr::extract(pixel, "pixel", "(\\d+)", convert = TRUE) |>
  mutate(pixel = pixel - 2, x = pixel %% 28, y = 28 - pixel %/% 28) |>
  filter(instance %in% c(4175, 3997, 5378, 387, 1854))

theme_set(theme_light())
imge_error_sample5 <- pixels_gathered |> 
  ggplot(aes(x, y, fill = value)) +  
  geom_tile() +  
  facet_wrap(~ instance, ncol = 5) +
  theme(legend.position = "none", plot.title = element_text(size = 7, hjust = 0.5, vjust = -0.5),
              axis.title.x = element_blank(), axis.title.y = element_blank(),
              axis.text.x = element_blank(), axis.ticks.x = element_blank(),
              axis.text.y = element_blank(), axis.ticks.y = element_blank(),
              panel.grid.major = element_blank(), panel.grid.minor = element_blank(), #change legend key width
        legend.title = element_text(size=5), #change legend title font size
        legend.text = element_text(size=4),
         legend.key.height = unit(0.25, 'cm'),
         legend.key.width = unit(0.25, 'cm'),
        strip.text = element_blank())
```

<!--images with position 6-->
```{r}
#| warning: false
#| echo: false

pixels_gathered <-  mnist_data |>
  mutate(instance = row_number()) |>
  gather(pixel, value, -Label, -instance) |>
  tidyr::extract(pixel, "pixel", "(\\d+)", convert = TRUE) |>
  mutate(pixel = pixel - 2, x = pixel %% 28, y = 28 - pixel %/% 28) |>
  filter(instance %in% c(614, 3079, 1762, 5239, 3723))

theme_set(theme_light())
imge_error_sample6 <- pixels_gathered |> 
  ggplot(aes(x, y, fill = value)) +  
  geom_tile() +  
  facet_wrap(~ instance, ncol = 5) +
  theme(legend.position = "none", plot.title = element_text(size = 7, hjust = 0.5, vjust = -0.5),
              axis.title.x = element_blank(), axis.title.y = element_blank(),
              axis.text.x = element_blank(), axis.ticks.x = element_blank(),
              axis.text.y = element_blank(), axis.ticks.y = element_blank(),
              panel.grid.major = element_blank(), panel.grid.minor = element_blank(), #change legend key width
        legend.title = element_text(size=5), #change legend title font size
        legend.text = element_text(size=4),
         legend.key.height = unit(0.25, 'cm'),
         legend.key.width = unit(0.25, 'cm'),
        strip.text = element_blank())
```

<!--images with position 7-->
```{r}
#| warning: false
#| echo: false

pixels_gathered <-  mnist_data |>
  mutate(instance = row_number()) |>
  gather(pixel, value, -Label, -instance) |>
  tidyr::extract(pixel, "pixel", "(\\d+)", convert = TRUE) |>
  mutate(pixel = pixel - 2, x = pixel %% 28, y = 28 - pixel %/% 28) |>
  filter(instance %in% c(5748, 728, 7419, 7794, 6233))

theme_set(theme_light())
imge_error_sample7 <- pixels_gathered |> 
  ggplot(aes(x, y, fill = value)) +  
  geom_tile() +  
  facet_wrap(~ instance, ncol = 5) +
  theme(legend.position = "none", plot.title = element_text(size = 7, hjust = 0.5, vjust = -0.5),
              axis.title.x = element_blank(), axis.title.y = element_blank(),
              axis.text.x = element_blank(), axis.ticks.x = element_blank(),
              axis.text.y = element_blank(), axis.ticks.y = element_blank(),
              panel.grid.major = element_blank(), panel.grid.minor = element_blank(), #change legend key width
        legend.title = element_text(size=5), #change legend title font size
        legend.text = element_text(size=4),
         legend.key.height = unit(0.25, 'cm'),
         legend.key.width = unit(0.25, 'cm'),
        strip.text = element_blank())
```

<!--images with position 8-->
```{r}
#| warning: false
#| echo: false

pixels_gathered <-  mnist_data |>
  mutate(instance = row_number()) |>
  gather(pixel, value, -Label, -instance) |>
  tidyr::extract(pixel, "pixel", "(\\d+)", convert = TRUE) |>
  mutate(pixel = pixel - 2, x = pixel %% 28, y = 28 - pixel %/% 28) |>
  filter(instance %in% c(33, 2485, 5998, 318, 1761))

theme_set(theme_light())
imge_error_sample8 <- pixels_gathered |> 
  ggplot(aes(x, y, fill = value)) +  
  geom_tile() +  
  facet_wrap(~ instance, ncol = 5) +
  theme(legend.position = "none", plot.title = element_text(size = 7, hjust = 0.5, vjust = -0.5),
              axis.title.x = element_blank(), axis.title.y = element_blank(),
              axis.text.x = element_blank(), axis.ticks.x = element_blank(),
              axis.text.y = element_blank(), axis.ticks.y = element_blank(),
              panel.grid.major = element_blank(), panel.grid.minor = element_blank(), #change legend key width
        legend.title = element_text(size=5), #change legend title font size
        legend.text = element_text(size=4),
         legend.key.height = unit(0.25, 'cm'),
         legend.key.width = unit(0.25, 'cm'),
        strip.text = element_blank())
```

<!--images with position 9-->
```{r}
#| warning: false
#| echo: false

pixels_gathered <-  mnist_data |>
  mutate(instance = row_number()) |>
  gather(pixel, value, -Label, -instance) |>
  tidyr::extract(pixel, "pixel", "(\\d+)", convert = TRUE) |>
  mutate(pixel = pixel - 2, x = pixel %% 28, y = 28 - pixel %/% 28) |>
  filter(instance %in% c(5690, 165, 517, 6935, 2682))

theme_set(theme_light())
imge_error_sample9 <- pixels_gathered |> 
  ggplot(aes(x, y, fill = value)) +  
  geom_tile() +  
  facet_wrap(~ instance, ncol = 5) +
  theme(legend.position = "none", plot.title = element_text(size = 7, hjust = 0.5, vjust = -0.5),
              axis.title.x = element_blank(), axis.title.y = element_blank(),
              axis.text.x = element_blank(), axis.ticks.x = element_blank(),
              axis.text.y = element_blank(), axis.ticks.y = element_blank(),
              panel.grid.major = element_blank(), panel.grid.minor = element_blank(), #change legend key width
        legend.title = element_text(size=5), #change legend title font size
        legend.text = element_text(size=4),
         legend.key.height = unit(0.25, 'cm'),
         legend.key.width = unit(0.25, 'cm'),
        strip.text = element_blank())
```

<!--images with position 10-->
```{r}
#| warning: false
#| echo: false

pixels_gathered <-  mnist_data |>
  mutate(instance = row_number()) |>
  gather(pixel, value, -Label, -instance) |>
  tidyr::extract(pixel, "pixel", "(\\d+)", convert = TRUE) |>
  mutate(pixel = pixel - 2, x = pixel %% 28, y = 28 - pixel %/% 28) |>
  filter(instance %in% c(4962, 2264, 5563, 6369, 559))

theme_set(theme_light())
imge_error_sample10 <- pixels_gathered |> 
  ggplot(aes(x, y, fill = value)) +  
  geom_tile() +  
  facet_wrap(~ instance, ncol = 5) +
  theme(legend.position = "none", plot.title = element_text(size = 7, hjust = 0.5, vjust = -0.5),
              axis.title.x = element_blank(), axis.title.y = element_blank(),
              axis.text.x = element_blank(), axis.ticks.x = element_blank(),
              axis.text.y = element_blank(), axis.ticks.y = element_blank(),
              panel.grid.major = element_blank(), panel.grid.minor = element_blank(), #change legend key width
        legend.title = element_text(size=5), #change legend title font size
        legend.text = element_text(size=4),
         legend.key.height = unit(0.25, 'cm'),
         legend.key.width = unit(0.25, 'cm'),
        strip.text = element_blank())
```

<!--images with position 11-->
```{r}
#| warning: false
#| echo: false

pixels_gathered <-  mnist_data |>
  mutate(instance = row_number()) |>
  gather(pixel, value, -Label, -instance) |>
  tidyr::extract(pixel, "pixel", "(\\d+)", convert = TRUE) |>
  mutate(pixel = pixel - 2, x = pixel %% 28, y = 28 - pixel %/% 28) |>
  filter(instance %in% c(5188, 4849, 2666, 3448, 3055))

theme_set(theme_light())
imge_error_sample11 <- pixels_gathered |> 
  ggplot(aes(x, y, fill = value)) +  
  geom_tile() +  
  facet_wrap(~ instance, ncol = 5) +
  theme(legend.position = "none", plot.title = element_text(size = 7, hjust = 0.5, vjust = -0.5),
              axis.title.x = element_blank(), axis.title.y = element_blank(),
              axis.text.x = element_blank(), axis.ticks.x = element_blank(),
              axis.text.y = element_blank(), axis.ticks.y = element_blank(),
              panel.grid.major = element_blank(), panel.grid.minor = element_blank(), #change legend key width
        legend.title = element_text(size=5), #change legend title font size
        legend.text = element_text(size=4),
         legend.key.height = unit(0.25, 'cm'),
         legend.key.width = unit(0.25, 'cm'),
        strip.text = element_blank())
```

<!--images with position 12-->
```{r}
#| warning: false
#| echo: false

pixels_gathered <-  mnist_data |>
  mutate(instance = row_number()) |>
  gather(pixel, value, -Label, -instance) |>
  tidyr::extract(pixel, "pixel", "(\\d+)", convert = TRUE) |>
  mutate(pixel = pixel - 2, x = pixel %% 28, y = 28 - pixel %/% 28) |>
  filter(instance %in% c(3120, 6869, 6345, 4470, 7147))

theme_set(theme_light())
imge_error_sample12 <- pixels_gathered |> 
  ggplot(aes(x, y, fill = value)) +  
  geom_tile() +  
  facet_wrap(~ instance, ncol = 5) +
  theme(legend.position = "none", plot.title = element_text(size = 7, hjust = 0.5, vjust = -0.5),
              axis.title.x = element_blank(), axis.title.y = element_blank(),
              axis.text.x = element_blank(), axis.ticks.x = element_blank(),
              axis.text.y = element_blank(), axis.ticks.y = element_blank(),
              panel.grid.major = element_blank(), panel.grid.minor = element_blank(), #change legend key width
        legend.title = element_text(size=5), #change legend title font size
        legend.text = element_text(size=4),
         legend.key.height = unit(0.25, 'cm'),
         legend.key.width = unit(0.25, 'cm'),
        strip.text = element_blank())
```

```{r}
#| echo: false
#| fig-cap: "Images of the handwritten digit 1 are distributed along the left-to-right of first embedding axis of the PaCMAP nonlinear structure."
#| label: fig-sampleimg
#| out-width: 100%
#| fig-pos: H

imge_error_sample1 + imge_error_sample2 + imge_error_sample3 + imge_error_sample4 + imge_error_sample5 + imge_error_sample6 + imge_error_sample7 + imge_error_sample8 + imge_error_sample9 + imge_error_sample10 + imge_error_sample11 + imge_error_sample12 +
  plot_annotation(tag_levels = 'a') +
  plot_layout(ncol = 2, nrow = 6) &
  theme(plot.tag = element_text(size = 8))
```

mse_df_mnist <- dplyr::bind_rows(vec)[0, ]
mse_df_mnist <- mse_df_mnist |>
  dplyr::mutate_if(is.character, as.numeric)

  num_bin_list <- calc_bins(data = pacmap_minst_scaled, 
            x = "PaCMAP1", y = "PaCMAP2", 
            hex_size = hex_size_vec[i], buffer_x = NA, buffer_y = NA)

  model_object <- fit_highd_model( training_data = training_data_mnist, 
                                   nldr_df_with_id = pacmap_minst_scaled, 
                                   x = "PaCMAP1", y = "PaCMAP2", 
                                   num_bins_x = num_bins_x, 
                                   num_bins_y = num_bins_y, 
                                   x_start = NA, y_start = NA, 
                                   buffer_x = NA, buffer_y = NA, 
                                   hex_size = hex_size_vec[i],
                                   is_rm_lwd_hex = FALSE, 
                                   benchmark_to_rm_lwd_hex = NA, 
                                   col_start_2d = "PaCMAP", 
                                   col_start_highd = "PC")

  pred_emb_list <- predict_emb(test_data = training_data_mnist, 
                                  df_bin_centroids = centroid_df_training, 
                                  df_bin = avg_df_training, type_NLDR = "PaCMAP")

  eval_list <- gen_summary(test_data = training_data_mnist, 
                                  prediction_df = pred_df_training, 
                                  df_bin = avg_df_training, col_start = "PC")

  mse_df_mnist <- mse_df_mnist |>
    tibble::add_row(num_bins = num_bins_x * num_bins_y,
                    aic = eval_list$aic,
                    mse = eval_list$mse,
                    num_bins_x = num_bins_x,
                    num_bins_y = num_bins_y,
                    hex_size = hex_size_vec[i])

## If same total number of bins occurred only select ones with minimum error
### Obtain duplicate bins
dupli_bins <- mse_df_mnist |> 
  dplyr::count(num_bins) |> 
  dplyr::filter(n > 1) |> 
  dplyr::pull(num_bins)

### Group split by duplicated bins
duplicate_df_list <- mse_df_mnist |>
  dplyr::filter(num_bins %in% dupli_bins) |>
  dplyr::arrange(num_bins) |>
  dplyr::group_split(num_bins)

### Obtain the mse_df with not duplicated bins
not_dupli_df <- mse_df_mnist |>
  dplyr::filter(!(num_bins %in% dupli_bins))

### Combine duplicated and not duplicated(corrected) bins dfs
mse_df_mnist <- dplyr::bind_rows(not_dupli_df, duplicate_df)
```

mse_plot_mnist <- ggplot(mse_df_mnist, aes(x = num_bins,
                                       y = mse
)) +
  geom_point() +
  geom_line() +
   geom_vline(xintercept = 180, linetype="solid",
                color = "red", size=0.8, alpha = 0.5) +
  theme_light() +
  theme(legend.position = "none", legend.title = element_blank(), plot.title = element_text(size = 7, hjust = 0.5, vjust = -0.5),
        axis.title = element_text(size = 7),
        axis.text = element_text(size = 7)) +
  #scale_colour_manual(values = c("#377eb8")) +
  ylab("MSE") +
  xlab("total number of bins")

```{r}
#| echo: false
#| fig-cap: "Goodness of fit statistics from PaCMAP applied to training MNIST dataset. What is the effective number of bins in PaCMAP to create a 2D model? The MSE plot have a steep slope at the beginning, indicating that a smaller number of bins causes a larger amount of error. Then, the slope gradually declines or level off, indicating that a higher number of bins generates a smaller error. Using the elbow method, when the total number of bins is set to 180, the slope of the Mean Squared Error (MSE) plot experiences a sudden and noticeable change, resembling an elbow-like shape. This point indicates that adding less bins does not enough to capture the data structure."
#| label: fig-diagnosticpltMNIST
#| fig-pos: H

mse_plot_mnist
```

@@

error_df <- error_df |>
  mutate(type = if_else(total == 0, "no error",
                        if_else(total <= 2, "error 0-2",
                                if_else(total <= 4, "error 2-4",
                                        if_else(total <= 6, "error 4-6",
                                                if_else(total <= 10, "error 6-10",
                                                        if_else(total <= 12, "error 10-12",
                                                                if_else(total <= 14, "error 12-14", "error greter than 14"))))))))

plot_list2_mnist_error <- error_df |>
  mutate(type = factor(type , levels = c("no error", "error 0-2", "error 2-4", "error 4-6", "error 6-10", "error 10-12", "error 12-14", "error greter than 14"))) |>
    ggplot(aes(x = PaCMAP1,
               y = PaCMAP2, color = type, group = ID))+
    geom_point(alpha=0.5, size = 0.1) +
    coord_equal() +
    theme(plot.title = element_text(hjust = 0.5, size = 18, face = "bold")) + #ggtitle("(a)") +
  theme_linedraw() +
    theme(plot.title = element_text(size = 7, hjust = 0.5, vjust = -0.5),
              axis.title.x = element_blank(), axis.title.y = element_blank(),
              axis.text.x = element_blank(), axis.ticks.x = element_blank(),
              axis.text.y = element_blank(), axis.ticks.y = element_blank(),
              panel.grid.major = element_blank(), panel.grid.minor = element_blank(), #change legend key width
        legend.title = element_text(size=5), #change legend title font size
        legend.text = element_text(size=4),
         legend.key.height = unit(0.25, 'cm'),
         legend.key.width = unit(0.25, 'cm')) +
  scale_color_manual(values=c("#fee0d2", "#fcbba1",
                                         "#fc9272", "#fb6a4a", "#ef3b2c",
                                         "#cb181d", "#a50f15", "#99000d")) +
  annotate(geom = 'text', label = 'b', x = Inf, y = Inf, hjust = 1.5, vjust = 1.5, size = 3)

@@

```{r}
#| echo: false
#| fig-cap: "(a) Model in the 2D space overlaid on PaCMAP data, and (b) high-D model error in model space."
#| label: fig-nldervisMNISTPACMAP
#| fig-pos: H

trimesh_removed_mnist_pacmap + plot_list2_mnist_error +
plot_layout(guides='collect', ncol=3) &
  theme(legend.position='bottom')
```

>>>>

plot_list2_pbmc_error <- error_df |>
  mutate(type = factor(type , levels = c("no error", "error 0-5", "error 5-10", "error 10-15", "error 15-20", "error 20-25", "error greter than 25"))) |>
    ggplot(aes(x = UMAP1,
               y = UMAP2, color = type))+
    geom_point(alpha=0.5, size = 0.1) +
    coord_equal() +
    theme(plot.title = element_text(hjust = 0.5, size = 18, face = "bold")) + #ggtitle("(a)") +
  theme_linedraw() +
    theme(plot.title = element_text(size = 7, hjust = 0.5, vjust = -0.5),
              axis.title.x = element_blank(), axis.title.y = element_blank(),
              axis.text.x = element_blank(), axis.ticks.x = element_blank(),
              axis.text.y = element_blank(), axis.ticks.y = element_blank(),
              panel.grid.major = element_blank(), panel.grid.minor = element_blank(), #change legend key width
        legend.title = element_text(size=5), #change legend title font size
        legend.text = element_text(size=4),
         legend.key.height = unit(0.25, 'cm'),
         legend.key.width = unit(0.25, 'cm')) +
  scale_color_manual(values=rainbow_hcl(7)) +
  annotate(geom = 'text', label = 'b', x = -Inf, y = Inf, hjust = -0.5, vjust = 1.5, size = 3)
```

@@

The MNIST dataset consists of grayscale images of handwritten digits [@lecun2010]. @Yingfan2021 used this dataset to demonstrate how the PaCMAP preserves local structure. We selected the 2D embedding of PaCMAP for the handwritten digit 1. As shown in @fig-pacmapauthor, we observe a nonlinear structure present in 2D space. Furthermore, the angle of digit 1 images varies along the nonlinear structure (see @fig-sampleimg). To answer the following questions,

@@

According to @fig-mnist1_sc1, the nonlinear continuity structure observed in the 2D representation of PaCMAP is also visible when visualizing the model overlaid on the data space. This provides evidence that the model accurately captures the structure of the high-D data. Also, the model has a twisted pattern within the nonlinear structure, which is an addition pattern of data that cannot be seen in the 2D representation (see @fig-mnist1_sc2).

In order to visually assess the preservation of local structure, it's important to consider the edges in the model. In a model that effectively preserves local structure, there shouldn't be any long edges in high-D space. However, as shown in @fig-mnist1_sc3, some long edges exist in high-D space that are not identified as long edges in the 2D representation.

@@

<!--images with positions-->
```{r}
#| warning: false
#| echo: false

pixels_gathered <-  mnist_data |>
  mutate(instance = row_number()) |>
  gather(pixel, value, -Label, -instance) |>
  tidyr::extract(pixel, "pixel", "(\\d+)", convert = TRUE) |>
  mutate(pixel = pixel - 2, x = pixel %% 28, y = 28 - pixel %/% 28) |>
  filter(instance %in% c(3710, 2391, 2385, 5030, 6475,
                         7679, 3568, 1312, 3311, 1097,
                         3552, 7853, 6489, 7689, 6690,
                        1380, 6057, 2347, 5946, 3355,
                        4175, 3997, 5378, 387, 1854,
                        614, 3079, 1762, 5239, 3723,
                        5748, 728, 7419, 7794, 6233,
                        33, 2485, 5998, 318, 1761,
                        5690, 165, 517, 6935, 2682,
                        4962, 2264, 5563, 6369, 559,
                        5188, 4849, 2666, 3448, 3055,
                        3120, 6869, 6345, 4470, 7147)) |>
  mutate(instance = as.character(instance))

theme_set(theme_light())
img_sample <- pixels_gathered |> 
  ggplot(aes(x, y, fill = value)) +  
  geom_tile() +  
  facet_wrap(~ factor(instance, levels = c(3710, 2391, 2385, 5030, 6475,
                         7679, 3568, 1312, 3311, 1097,
                         3552, 7853, 6489, 7689, 6690,
                        1380, 6057, 2347, 5946, 3355,
                        4175, 3997, 5378, 387, 1854,
                        614, 3079, 1762, 5239, 3723,
                        5748, 728, 7419, 7794, 6233,
                        33, 2485, 5998, 318, 1761,
                        5690, 165, 517, 6935, 2682,
                        4962, 2264, 5563, 6369, 559,
                        5188, 4849, 2666, 3448, 3055,
                        3120, 6869, 6345, 4470, 7147)), ncol = 10) +
  theme(legend.position = "none", plot.title = element_text(size = 7, hjust = 0.5, vjust = -0.5),
              axis.title.x = element_blank(), axis.title.y = element_blank(),
              axis.text.x = element_blank(), axis.ticks.x = element_blank(),
              axis.text.y = element_blank(), axis.ticks.y = element_blank(),
              panel.grid.major = element_blank(), panel.grid.minor = element_blank(), #change legend key width
        legend.title = element_text(size=5), #change legend title font size
        legend.text = element_text(size=4),
         legend.key.height = unit(0.25, 'cm'),
         legend.key.width = unit(0.25, 'cm'),
        strip.text = element_blank(),
        aspect.ratio = 1)
```

```{r}
#| echo: false
#| fig-cap: "Images of the handwritten digit 1 are distributed along the right-to-left of first embedding axis of the PaCMAP nonlinear structure."
#| label: fig-sampleimg
#| out-width: 100%
#| fig-pos: H

img_sample
```

@@

error_df <- error_df |>
  mutate(type = if_else(total <= 2, "error 0-2",
                        if_else(total <= 4, "error 2-4",
                                if_else(total <= 6, "error 4-6",
                                        if_else(total <= 8, "error 6-8",
                                                if_else(total <= 10, "error 8-10",
                                                        if_else(total <= 12, "error 10-12",
                                                                if_else(total <= 14, "error 12-14", "error greter than 14"))))))))

plot_list2_mnist_error <- error_df |>
  mutate(type = factor(type , levels = c("error 0-2", "error 2-4", "error 4-6", "error 6-8", "error 8-10", "error 10-12", "error 12-14", "error greter than 14"))) |>
    ggplot(aes(x = PaCMAP1,
               y = PaCMAP2, color = type, group = ID))+
    geom_point(alpha=0.5, size = 0.1) +
    coord_equal() +
    theme(plot.title = element_text(hjust = 0.5, size = 18, face = "bold")) + #ggtitle("(a)") +
  theme_linedraw() +
    theme(plot.title = element_text(size = 7, hjust = 0.5, vjust = -0.5),
              axis.title.x = element_blank(), axis.title.y = element_blank(),
              axis.text.x = element_blank(), axis.ticks.x = element_blank(),
              axis.text.y = element_blank(), axis.ticks.y = element_blank(),
              panel.grid.major = element_blank(), panel.grid.minor = element_blank(), #change legend key width
        legend.title = element_text(size=5), #change legend title font size
        legend.text = element_text(size=4),
         legend.key.height = unit(0.25, 'cm'),
         legend.key.width = unit(0.25, 'cm')) +
  scale_color_manual(values=rainbow_hcl(8)) +
  annotate(geom = 'text', label = 'b', x = Inf, y = Inf, hjust = 1.5, vjust = 1.5, size = 3)

@@

```{r}
#| echo: false
#| fig-cap: "(a) Model generated with 12 and 15 bins along x and y axes with hexagonal size 0.06 in the 2D space overlaid on PaCMAP data, and (b) high-D model error in model space."
#| label: fig-nldervisMNISTPACMAP
#| fig-pos: H

trimesh_removed_mnist_pacmap + plot_list2_mnist_error +
plot_layout(guides='collect', ncol=2) &
  theme(legend.position='bottom')
```

::: {#fig-mnist1_sc layout-ncol="3" fig-pos="H"}
![](figures/mnist/sc_1.png){#fig-mnist1_sc1}

![](figures/mnist/sc_2.png){#fig-mnist1_sc2}

![](figures/mnist/sc_3.png){#fig-mnist1_sc3}

Screen shots of the **langevitour** of the MNIST digit 1 data set, shows the model-in-data space, a video of the tour animation is available at (<https://youtu.be/zq2GM9qvUNA>).
:::

In terms of model error, some data points show high error (see @fig-nldervisMNISTPACMAP) due to their deviation from the typical high-D data structure, acting as outliers. This is evident when comparing the images associated with these high model error points (see @fig-sampleimgerror) to others (see @fig-sampleimg).

<!--images that occur large error-->
```{r}
#| warning: false
#| echo: false

img_error <- error_df |> 
  dplyr::filter(total >= 10) |> 
  dplyr::pull(ID)

mnist_data <- read_rds("data/mnist/mnist_digit_1.rds")
pixels_gathered <-  mnist_data |>
  mutate(instance = row_number()) |>
  gather(pixel, value, -Label, -instance) |>
  tidyr::extract(pixel, "pixel", "(\\d+)", convert = TRUE) |>
  mutate(pixel = pixel - 2, x = pixel %% 28, y = 28 - pixel %/% 28) |>
  filter(instance %in% img_error)

theme_set(theme_light())
imge_error_sample <- pixels_gathered |> 
  ggplot(aes(x, y, fill = value)) +  
  geom_tile() +  
  facet_wrap(~ instance, ncol = 11) +
  theme(legend.position = "none", plot.title = element_text(size = 7, hjust = 0.5, vjust = -0.5),
              axis.title.x = element_blank(), axis.title.y = element_blank(),
              axis.text.x = element_blank(), axis.ticks.x = element_blank(),
              axis.text.y = element_blank(), axis.ticks.y = element_blank(),
              panel.grid.major = element_blank(), panel.grid.minor = element_blank(), #change legend key width
        legend.title = element_text(size=5), #change legend title font size
        legend.text = element_text(size=4),
         legend.key.height = unit(0.25, 'cm'),
         legend.key.width = unit(0.25, 'cm'),
        strip.text = element_blank(),
        aspect.ratio = 1)
```

```{r}
#| echo: false
#| fig-cap: "Images of handwritten digit 1 which occur large model error."
#| label: fig-sampleimgerror
#| out-width: 100%
#| fig-pos: H

imge_error_sample
```
